\Large Generation of Mathematical Programming Representation for Discrete Event Simulation Models of Timed Petri Nets

### 2023-01-01 Answer to reviewers.tex

The usefulness %and benefit 
of this work can be justified in two aspects, namely generalization and convenience. TPN is a widely--used general modeling diagram of DEVS, and the proposed work would be easy to apply for people in this field.

The computational complexity of using MP to optimize DEVS has been %noticed 
addressed in many works, and they dealt the issue with classic MP techniques such as Benders decomposition, linear approximation, columns-row generation. We cannot deny that the computation issue really exists, but the ways to overcome are also vast. Furthermore, the translation is bidirectional. One can translate simulation into MP to optimize a system, and also translate the resulting MP into simulation. The value of translating MP into simulation is that the research on MP (such as duality and sensitivity) can %dig
increase the value of simulation, which is currently usually regarded only as a performance evaluation tool.

In the following, we answer the comments in details.

\section{DE comments}
\textit{The merit of the proposed technique should be clarified either in a general setting or on specific classes of systems like Petri nets or finite automata, and comparisons with extant tools and algorithms should be made.}

\noindent
\textbf{Answer:} %In this submission, 
In the revised manuscript our approach %works
has been reformulated to have it working under general settings deterministic timed Petri net. With an example of the $G/G/m$ queue, the differences between the proposed model and the state-of-the-art model are discussed.

@@

The benefits of using MP model to represent discrete event systems are many. As already mentioned above, when coupled with optimization, the MP-based algorithms can be faster and %gain 
reach better solution than black--box optimization algorithms. Furthermore, black--box approaches have limited capability is solving constrained optimization problems, %but 
while MP-based approaches can easily deal with them (\cite{zhang2020models}). Third, the vast theoretical and methodological results developed in the MP field can be introduced into the study of DEVS through simulation. For instance, using sensitivity analysis of MP models, DES can %be
become a gradient estimation tool, as suggested in \cite{chan2008optimization}. In fact, we do not proposed to totally replace simulation with MP, but to enrich the toolbox for analyzing and optimizing DEVS. Finally, the application of MP models of DEVS is not limited to optimization. \cite{basile2012k} proposed the sufficient and necessary condition of K-diagnosability of TPN based on MP.

The major concern about the application of MP is the computational complexity, %and the solving procedure 
as the solution procedure can be time-consuming or even unbearable. However, many approaches from optimization community are available to improve the efficiency. Linear programming approximation (\cite{alfieri2012mathematical}), Benders decomposition (\cite{weiss2015buffer}), row-column generation (\cite{alfieri2020time}), have been studied to solve optimization problems in %reality
real contexts based-on MP model of DEVS.

~\\

\textit{2. Furthermore, the discussion about the application of the approach should be significantly strengthened. Currently, this is rather vague and there is no comparisons with existing approaches that are used to address these problems.}

\noindent
\textbf{Answer: With an example of the $G/G/m$ queue, the differences between the proposed model and the state-of-the-art model are discussed.}

@@

\textit{References: \cite{basile2012k}, \cite{basile2012k}, \cite{bemporad1999control}}

\noindent
\textbf{Answer:} As the reviewer suggests, %this submission proposed 
in the revised manuscript we propose the translation of TPN into MP.

The difference between the proposed approach and the suggested references are %as follows.
detailed in the following. \cite{bemporad1999control} proposed a mixed integer programming modeling framework for hybrid systems, whose states are mixed integer, in discrete time, while this work deals with continuous--time discrete--state systems. An MP model was proposed to minimize the makespan of single-server manufacturing systems in \cite{di2020makespan}, %but this
while our work is more general and can model multi-server systems. \cite{basile2012k} proposed the model based on PN, not TPN, because %it deals with 
the paper is about diagnosability %other 
rather than time-dependent performance measures, such as waiting time or throughput.

The effective application domain of the proposed technique is basically related to optimization. The MP model of TPN %is easy to expand 
can be easily expanded to an MP model of TPN optimization. Based on the proposed MP model, an example with the extended MP model to optimize the $G/G/m$ queue is presented in Section 4.3.

@@

\noindent
\textbf{Answer:}

~\\

\textit{2. Detailed Comments}

\textit{The paper reads clearly. There are only very few typos (see at the end).}
\textbf{Answer: we have correct the typos.}

\textit{From a technical point of view, there is one thing that needs to be clarified, page 8: How do you choose constant M? Understandably, it should be large enough. However, I would expect to find in the paper a discussion about a lower bound on M.}

\noindent
\textbf{Answer: In the subsections of Section 3, the value of M is discussed each time it appears.}

\textbf{Section 3.1: The value of M in constraints (A1) and (A2) can be set to a value that is larger than the simulation time span.}

\textbf{Section 3.2: The value of M can be set to a value that is larger than the simulation time span for constraints (B1) and (B2).  The value of $M$ can be chosen as the upper bound of marking of place $p$ minus $(W^{p,t}-1)$ for constraints (B4). }

\textbf{Section 3.3: The value of $M$ in constraints (C3) and (C4) can be set to $K$. }

### 2023-01-01 main.tex
mengyi's revision

Discrete event dynamic systems (DEVS) have wide application in manufacturing and service applications. Even though many theoretical and analytical studies on discrete event dynamic systems have been developed, performance of real systems with high degree of complexity can only be evaluated with discrete event simulation (DES). Hence, simulation--optimization algorithms are widely used when numerical performance evaluation must be coupled with optimization, i.e., when the best system configuration, according to some criteria, has to be found meanwhile guaranteeing a given value of some performance measures \citep{fu20115book}.

@@

The rest of the paper is organized as follows. Section \ref{sec:sim} briefly introduces the TPN and its simulation algorithm. Section \ref{sec:TPN_MPR} describes the generation of the MP model from an existing TPN model. Section XXX shows several examples of application and validation of the equivalence between simulation and the MP. Discussion and conclusion are reported in Section XXX and XXX, respectively.

@@

\subsection{Basic definitions}
A TPN is a directed bipartite graph. The set of nodes in that graph is partitioned into a set of \textit{places} $P$ and a set of \textit{transitions} $T$. Places and transitions are connected with weighted and directed edges. In the example shown in the right side of Figure \ref{fig:gg1_petri}, places (i.e., $p_1$, $p_2$, $p_3$), transitions (i.e., $t_1$, $t_2$) and tokens are represented by white circles, rectangles and small black circles, respectively. Places hold \textit{tokens}, and tokens are transferred through transition \textit{firings}, i.e., one transition firing absorbs tokens from some places and releases them to some other places. Each transition $t\in{T}$ can absorb tokens from all of its precedents, and the number of tokens absorbed from place $p\in P$ is equal to the weight of the edge connecting $p$ and $t$. The number of tokens in each place must be non--negative at any time; hence, a firing of transition $t$ is enabled if and only if the number of tokens in all its precedents $p$ is not smaller than the weight on edge $(p,t)$. Each transition $t\in{T}$ can release tokens to all of its successors, and the number of tokens released to place $p\in P$ is equal to the weight of the edge connecting $t$ and $p$.  It is not necessary that the set of precedents and the set of successors are mutually exclusive. 
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{Figures/GG1_petri.png}
	\caption{G/G/2 queue and its TPN model.}
	\label{fig:gg1_petri}
\end{figure}
In a TPN, the duration between the moment of absorption and the moment of release is not always negligible, and the duration is called \textit{firing time}. This work considers firing time %non-negative and that
that are non--negative and can follow arbitrary stochastic distribution.

@@

(\textbf{Arianna: First of all, sets $\mathbb{K}$, $\mathbb{T}$ and $\mathbb{I}^t$ are used to indicate the set of iterations, the set of transitions and the set of firings of transition $t$, respectively.})
Binary variables $w^{t}_{i,k}$ are introduced to represent that the $i$-th start firing of transition $t$ finishes in iteration $k$. If $w^{\xi}_{i,k}$ (\textbf{Andrea: this should be $w^{t}_{i,k}$}) is equal to one, %and 
variable $e^{t,f}_{i}$ is bounded to $\mathcal{E}_k$, as shown in constraints (A1) and (A2). \textbf{Andrea: The set $\mathbb{K}$ is not defined as well as the other sets. Arianna: I have tried to define them above. Mengyi please check} Constraints (A3) state that each firing-finish event executes at most once. Constraints (A4) state that, in each iteration, exactly one firing-finish event is executed. Constraints (A5) imply that the clock cannot be reversed from iteration $k-1$ to iteration $k$. Constraint (A6) implies that the simulation clock is initialized to zero. Constraints (A7) implies that the delay between the start and finish of firing transition $t$ is equal to a sample from the random variable $T^{t}$.

\begin{eqnarray}
	e^{t,f}_i-\mathcal{E}_k\ge M(w^{t}_{i,k}-1) & \forall\ t\in\mathbb{T},i\in \mathbb{I}^{t},k\in \mathbb{K}&(A1)\nonumber\\
	\mathcal{E}_k-e^{t,f}_i\ge M(w^{t}_{i,k}-1) & \forall\ t\in\mathbb{T},i\in \mathbb{I}^{t},k\in \mathbb{K}&(A2)\nonumber\\
	\sum_{k\in \mathbb{K}} w^{t}_{i,k} \le 1& \forall\ t\in\mathbb{T},i\in \mathbb{I}^{t}&(A3)\nonumber\\
	\sum_{t\in \mathbb{T}}\sum_{i\in \mathbb{I}^{t}} w^{t}_{i,k} =1&\forall\ k\in \mathbb{K}&(A4)\nonumber\\
	\mathcal{E}_{k}-\mathcal{E}_{k-1}\ge 0&\forall\ k\in \mathbb{K}&(A5)\nonumber\\
	\mathcal{E}_0 = 0&&(A6)\nonumber\\
	e^{t,f}_{i} - e^{t,s}_{i} = \tau^{t}_{i} & \forall t\in\mathbb{T}, i\in \mathbb{I}^{t}&(A7) \nonumber
\end{eqnarray}

@@

The second group of constraints, denoted by group--B, depicts the start of transition firing, as in lines 5 to 13 in Algorithm 1. Binary variables $x^{t}_{i,k}$ represent that a fire of transition $t$ starts in iteration $k$ if it is equal to one. Constraints (B1) and (B2) shows that the start firing time is equal to the simulation clock.

\begin{eqnarray}
	e^{t,s}_i-\mathcal{E}_{k} \ge M(x^{t}_{i,k}-1)& \forall\ t\in \mathbb{T},k\in \mathbb{K},i\in \mathbb{I}^{t}&(B1)\nonumber\\
	\mathcal{E}_{k} -e^{t,s}_i\ge M(x^{t}_{i,k}-1)&\forall\ t\in \mathbb{T},k\in \mathbb{K},i\in \mathbb{I}^{t}&(B2)\nonumber
\end{eqnarray}

The condition to start firing transition $t$ is that the tokens in all the precedent places $p$ are above the required level $A^{p,t}$ (\textbf{Andrea: this should be $W^{p,t}$}?). Binary variables $z^{t}_{k}$ equal to one represents the condition for transition $t$ is true, as constraints (B3). Moreover, a set of binary variables $v^{t,p}_k$ is used to verify if the number of tokens in precedent place $p$ is smaller than $W^{p,t}$, as constraints (B4) (\textbf{Andrea: $v^{t,p}_k$ is not well explained. It is equal to 1 if ...?}). Constraints (B5) assures that $z^{t}_k$ is equal to zero only if at least one precedent place does not contain enough number of tokens. Constraints (B6) show that if $z^{t}_k$ is equal to one, transition $t$ must start to fire in iteration $k$. Constraints (B7) state that at most one firing of any transition starts for each transition.

\begin{eqnarray}
	m^p_k - W^{p,t} \ge M(z^{t}_{k}-1)&\forall\ t\in \mathbb{T},k\in \mathbb{K},p\in \mathbb{P}&(B3)\nonumber\\
	(W^{p,t}-1) - m^p_k \ge M(v^{t,p}_k-1) & \forall\ t\in \mathbb{T},k\in \mathbb{K},p\in \mathbb{P} &(B4)\nonumber\\
	1 - z^{t}_{k} \le \sum_{p\in \mathbb{P}} v^{p,t}_k&\forall\ t\in \mathbb{T},k\in \mathbb{K}&(B5)\nonumber\\
	\sum_{i\in \mathbb{I}^{t}} x^{t}_{i,k} = z^{t}_k&\forall\ t\in \mathbb{T},k\in \mathbb{K}&(B6)\nonumber\\
	\sum_{k\in \mathbb{K}} x^{t}_{i,k} \le 1& \forall\ t\in \mathbb{T},i\in \mathbb{I}^{t}&(B7)\nonumber
\end{eqnarray}

@@

\subsection{Event sequencing}
The index $i$ of event $e^{t,s}_i$ represents the sequence of firing starting, i.e., if a firing starts in an earlier iteration, then its index will be smaller. Moreover, a firing must finish after starting. Group--C constraints force such sequences, which are relevant to all the events. Constraints (C1) show that if the $i$--th firing of transition $t$ does not start before simulation termination, then the $(i+1)$--th firing will not start. Constraints (C2) state that if the $i$--th firing of transition $t$ does not start before simulation termination, then it will not finish. Constraints (C3) state that a firing cannot finish before starting, unless it remains in the future event list at the end of the simulation run. Constraints (C4) depict that the $(i+1)$--th firing of transition $t$ must be scheduled after the $i$--th execution, unless the $(i+1)$--th execution is not scheduled before simulation termination.
\begin{eqnarray}
	\sum_{k\in \mathbb{K}} x^{t}_{i+1,k} - \sum_{k\in \mathbb{K}} x^{t}_{i,k} \le 0&  \forall\ t\in \mathbb{T},i\in \mathbb{I}^{t}&(C1)\nonumber\\
	\sum_{k\in \mathbb{K}} w^{t}_{i,k}-\sum_{k\in \mathbb{K}} x^{t}_{i,k} \le 0&  \forall\ t\in \mathbb{T},i\in \mathbb{I}^{t}&(C2)\nonumber\\
	\sum_{k\in \mathbb{K}} kw^{t}_{i,k} - \sum_{k\in \mathbb{K}} kx^{t}_{i,k} \ge M(\sum_{k\in \mathbb{K}} w^{t}_{i,k}-1) &\forall\ t\in \mathbb{T}, i\in \mathbb{I}^{t}&(C3) \nonumber\\
	\sum_{k\in \mathbb{K}} kx^{t}_{i+1,k} - \sum_{k\in \mathbb{K}} kx^{t}_{i,k} \ge 1+M(\sum_{k\in \mathbb{K}} x^{t}_{i+1,k}-1)&  \forall\ t\in \mathbb{T},i\in \mathbb{I}^{t}&(C4)\nonumber
\end{eqnarray}

@@

The tokens in place $p$ are changed from $m^{p}_{k}$ to $m^{p}_{k+1}$ in iteration $k$, as in constraints (E1) and in lines 16 to 18 in Algorithm 1. The transitions $t$ that start in %fire 
firing in iteration $k$ absorbs $A^{p,t}$ tokens from the precedent places $p$, and the firing that finishes in iteration $k$ releases tokens to the successive places.

\begin{eqnarray}
	m^p_{k+1} = m^p_k - \sum_{t\in \mathbb{T}} W^{p,t} \sum_{i\in \mathbb{I}^{t}} x^t_{i,k} +  \sum_{t\in \mathbb{T}} W^{t,p} \sum_{i\in \mathbb{I}^{t}} w^t_{i,k} &\forall p\in \mathbb{P}, k\in \mathbb{K}& (E1) \nonumber
\end{eqnarray}

\subsection{Objective function}\label{sec:TPN_obj}
With the constraints defined in the previous sections, there is a unique feasible solution in terms of transition firing times and the %history of simulation clock, 
sequence of simulation clock values, i.e., $e^{t,s}_i$, $e^{t,f}_{i}$ and $\mathcal{E}_k$. Thus, the objective function can be any function of %those
these variables, %for instance, 
such as, average %of 
system time or average waiting time in queueing systems. Notice that multiple feasible solutions may appear in terms of binary variables, since the sequence of events with identical execution time is not uniquely defined.

@@

\section{Applications}
In this section, several examples are presented, including a G/G/m queue, a merge queueing system composed by three single server stations, and a single server queue with failure as an example of event cancellation. The equivalence of MPR solution and history of a simulation run has been validated with $K$ equal to 20 for 100 independent replicates.

\subsection{G/G/m queue}\label{sec:TPN_ggm}
The first example is a G/G/m queue, and the TPN is shown in Figure \ref{fig:tpn_GG2}. In a G/G/m queue, customers arrive at the queue following an general arrival process. One of the identical idle servers processes customers in the queue one by one, and the processing time is generally distributed. After processing, the customer will be released from the system immediately, and the server becomes idle again. In Figure \ref{fig:tpn_GG2}, the TPN of the TPN of G/G/2 is presented. A token in place $p_{arr}$ shows that it is possible to schedule the arrival of next customer, and the firing time of transition $t_{arr}$ is equal to the inter-arrival time. At the end of firing $t_{arr}$, a customer arrives, and it is possible to schedule the next arrival, thus, it releases one token to $p_{arr}$. Furthermore, an arrival will increase the length of the queue by 1, which is represented by the number of tokens in place $p_{queue}$. Two tokens in place $p_{idle}$ shows that two servers are idle. The firing of transition $t_{process}$ shows the processing of a customer, and it is enabled if and only if at least one server is idle and there is a customer in the queue. Once the process of the customer finishes, i.e., the end of firing $t_{process}$, a server is released, and a token is sent to place $p_{idle}$.

@@

As mentioned in Section \ref{sec:MPR_start_Firing}, the transitions with non-zero firing times are expanded before implementing the MPR generation. The expanded TPN of the G/G/2 queue is shown in Figure \ref{fig:tpn_GG2_expansion}. Transition $t_{process}$ is expanded to $\tilde{t}_{process}$, $\bar{t}_{process}$ and $p_{process}$. Even though transition $t_{arr}$ also has non-zero firing time, it is self-limiting, i.e., there is never more than one firing in process. Therefore, it is not necessary to expand transition $t_{arr}$.

@@

The MPR of G/G/m queue is presented in the follows. The set of transitions $\mathbb{T}$ is specified by $\{t_arr, t_queue,\tilde{t}_{process}, \bar{t}_{process}\}$ for G/G/m queue. The constraints

Table \ref{tab:sim_GGM} shows a simulation run of a G/G/m queue including 10 executions. In iteration zero, the system is in initial state, and this state satisfies the condition to schedule the counting event of arrival, as in lines 2 to 8 in Algorithm 3. In the solution of MPR, the scheduling of $\tilde{arr}$ is represented by $z^{\tilde{arr}}_0$ and $x^{\tilde{arr}}_{1,0}$ equal to 1, and its occurring time is equal to the simulation clock, which is equal to 0. The execution $(\tilde{arr}, 1, e^{\tilde{arr},0}_1, e^{\tilde{arr},1}_1)$, denoted by $e^{\tilde{arr}}_1$ in the table, is then performed since it is the earliest execution (line 18 in Algorithm 3), which is also indicated by $w^{\tilde{arr}}_{1,1}$ equal to one, and simulation clock in the next iteration, i.e., iteration $1$, is equal to the occurring time $e^{\tilde{arr}}_{1,1}$ in the solution of MPR. After that, the positive-delay event $arr$ is then scheduled (lines 26 to 30 in Algorithm 3), which can also be seen from the MPR solution, since $x^{arr}_{1,1}$ is equal to one, the scheduling time of $e^{arr}_{1,0}$ is equal to the current simulation clock time and the occurrence time is 2.3 time unit later than the scheduling time, where 2.3 is sampled from the distribution of inter-arrival time $T^{arr}$. At the end of the iteration, the system state is updated according to the state transition function of event $\tilde{arr}$ (line 25 in Algorithm 3), which in the MPR solution is indicated by an increment equal to one of state variable $u^{arr}$. This procedure is repeated in the next iterations, and the equivalence of DES procedure and the solution of MPR can be seen from Table \ref{tab:sol_GGM}.

>>>>

Discrete event dynamic systems (DEVS) have wide application in manufacturing and service applications. Even though many theoretical and analytical studies on discrete event dynamic systems have been developed, performance of real systems with high degree of complexity can, sometimes, only be evaluated with discrete event simulation (DES). Hence, simulation--optimization algorithms are widely used when numerical performance evaluation must be coupled with optimization, i.e., when the best system configuration, according to some criteria, has to be found meanwhile guaranteeing a given value of some performance measures \citep{fu20115book}.

@@

The rest of the paper is organized as follows. Section \ref{sec:sim} briefly introduces the TPN and its simulation algorithm. Section \ref{sec:TPN_MPR} describes the generation of the MP model from an existing TPN model. Section \ref{sec:example} shows examples of application and validation of the equivalence between simulation and the MP. Discussion and conclusion are reported in Section\ref{sec:conclude}.

@@

\subsection{Basic definitions and notations}
A TPN is a directed bipartite graph. The set of nodes in that graph is partitioned into a set of \textit{places} $P$ and a set of \textit{transitions} $T$. Places and transitions are connected with weighted and directed edges. In the example shown in the right side of Figure \ref{fig:gg1_petri}, places (i.e., $p_1$, $p_2$, $p_3$), transitions (i.e., $t_1$, $t_2$) and tokens are represented by white circles, rectangles and small black circles, respectively. Places hold \textit{tokens}, and tokens are transferred through transition \textit{firings}, i.e., one transition firing absorbs tokens from some places and releases them to some other places. Each transition $t\in{T}$ can absorb tokens from all of its precedents, and the number of tokens absorbed from place $p\in P$ is equal to the weight of the edge connecting $p$ and $t$. The number of tokens in each place must be non--negative at any time; hence, a firing of transition $t$ is enabled if and only if the number of tokens in all its precedents $p$ is not smaller than the weight on edge $(p,t)$. Each transition $t\in{T}$ can release tokens to all of its successors, and the number of tokens released to place $p\in P$ is equal to the weight of the edge connecting $t$ and $p$.  It is not necessary that the set of precedents and the set of successors are mutually exclusive. 
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{Figures/GG1_petri.png}
	\caption{G/G/2 queue and its TPN model.}
	\label{fig:gg1_petri}
\end{figure}
In a TPN, the duration between the moment of absorption and the moment of release is not always negligible, and the duration is called \textit{firing time}. This work considers firing time %non-negative and that
that are non--negative and can follow arbitrary stochastic distribution.

@@

First of all, sets $\mathbb{K}$, $\mathbb{T}$ and $\mathbb{I}^t$ are used to indicate the set of iterations, the set of transitions and the set of firings of transition $t$, respectively.
Binary variables $w^{t}_{i,k}$ are introduced to represent that the $i$-th start firing of transition $t$ finishes in iteration $k$. If $w^{t}_{i,k}$  is equal to one, %and 
variable $e^{t,f}_{i}$ is bounded to $\mathcal{E}_k$, as shown in constraints (A1) and (A2). Constraints (A3) state that each firing-finish event executes at most once. Constraints (A4) state that, in each iteration, exactly one firing-finish event is executed. Constraints (A5) implies that the delay between the start and finish of firing transition $t$ is equal to a sample from the random variable $T^{t}$. Constraints (A6) imply that the clock cannot be reversed from iteration $k-1$ to iteration $k$. Constraint (A7) implies that the simulation clock is initialized to zero. The value of $M$ in constraints (A1) and (A2) can be set to a value that is larger than the simulation time span.

\begin{eqnarray}
	e^{t,f}_i-\mathcal{E}_k\ge M(w^{t}_{i,k}-1) & \forall\ t\in\mathbb{T},i\in \mathbb{I}^{t},k\in \mathbb{K}&(A1)\nonumber\\
	\mathcal{E}_k-e^{t,f}_i\ge M(w^{t}_{i,k}-1) & \forall\ t\in\mathbb{T},i\in \mathbb{I}^{t},k\in \mathbb{K}&(A2)\nonumber\\
	\sum_{k\in \mathbb{K}} w^{t}_{i,k} \le 1& \forall\ t\in\mathbb{T},i\in \mathbb{I}^{t}&(A3)\nonumber\\
	\sum_{t\in \mathbb{T}}\sum_{i\in \mathbb{I}^{t}} w^{t}_{i,k} =1&\forall\ k\in \mathbb{K}&(A4)\nonumber\\
	e^{t,f}_{i} - e^{t,s}_{i} = \tau^{t}_{i} & \forall t\in\mathbb{T}, i\in \mathbb{I}^{t}&(A5) \nonumber\\
	\mathcal{E}_{k}-\mathcal{E}_{k-1}\ge 0&\forall\ k\in \mathbb{K}&(A6)\nonumber\\
	\mathcal{E}_0 = 0&&(A7)\nonumber
\end{eqnarray}

@@

The second group of constraints, denoted by group--B, depicts the start of transition firing, as in lines 5 to 13 in Algorithm 1. Binary variables $x^{t}_{i,k}$ represent that a fire of transition $t$ starts in iteration $k$ if it is equal to one. Constraints (B1) and (B2) shows that the start firing time is equal to the simulation clock. The value of $M$ can be set to a value that is larger than the simulation time span.

\begin{eqnarray}
	e^{t,s}_i-\mathcal{E}_{k} \ge M(x^{t}_{i,k}-1)& \forall\ t\in \mathbb{T},k\in \mathbb{K},i\in \mathbb{I}^{t}&(B1)\nonumber\\
	\mathcal{E}_{k} -e^{t,s}_i\ge M(x^{t}_{i,k}-1)&\forall\ t\in \mathbb{T},k\in \mathbb{K},i\in \mathbb{I}^{t}&(B2)\nonumber%\\
	%e^{t,s}_i - e^{t,s}_{i-1} \ge 0&\forall\ t\in \mathbb{T},i\in \mathbb{I}^{t} &(B3)\nonumber
\end{eqnarray}

The condition to start firing transition $t$ is that the tokens in all the precedent places $p$ are above the required level $W^p,t$. Binary variables $z^{t}_{k}$ equal to one represents the condition for transition $t$ is true, as constraints (B3). Moreover, a set of binary variables $v^{t,p}_k$ is used to verify if the number of tokens in precedent place $p$ is smaller than $W^{p,t}$, as constraints (B4). The value of $M$ can be chosen as the upper bound of marking of place $p$ minus $(W^{p,t}-1)$. Variable $v^{t,p}_k$ is equal to 1, if marking in place $p$ is smaller than $W^{p,t}$. Constraints (B5) assures that $z^{t}_k$ is equal to zero only if at least one precedent place does not contain enough number of tokens. Constraints (B6) show that if $z^{t}_k$ is equal to one, transition $t$ must start to fire in iteration $k$. Constraints (B7) state that at most one firing of any transition starts for each transition.

\begin{eqnarray}
	m^p_k - W^{p,t} \ge W^{p,t}(z^{t}_{k}-1)&\forall\ t\in \mathbb{T},k\in \mathbb{K},p\in \mathbb{P}&(B3)\nonumber\\
	(W^{p,t}-1) - m^p_k \ge M(v^{t,p}_k-1) & \forall\ t\in \mathbb{T},k\in \mathbb{K},p\in \mathbb{P} &(B4)\nonumber\\
	1 - z^{t}_{k} \le \sum_{p\in \mathbb{P}} v^{p,t}_k&\forall\ t\in \mathbb{T},k\in \mathbb{K}&(B5)\nonumber\\
	\sum_{i\in \mathbb{I}^{t}} x^{t}_{i,k} = z^{t}_k&\forall\ t\in \mathbb{T},k\in \mathbb{K}&(B6)\nonumber\\
	\sum_{k\in \mathbb{K}} x^{t}_{i,k} \le 1& \forall\ t\in \mathbb{T},i\in \mathbb{I}^{t}&(B7)\nonumber
\end{eqnarray}

@@

\subsection{Event sequencing}
The index $i$ of event $e^{t,s}_i$ represents the sequence of firing starting, i.e., if a firing starts in an earlier iteration, then its index will be smaller. Moreover, a firing must finish after starting. Group--C constraints force such sequences, which are relevant to all the events. Constraints (C1) show that if the $i$--th firing of transition $t$ does not start before simulation termination, then the $(i+1)$--th firing will not start. Constraints (C2) state that if the $i$--th firing of transition $t$ does not start before simulation termination, then it will not finish. Constraints (C3) state that a firing cannot finish before starting, unless it remains in the future event list at the end of the simulation run. Constraints (C4) depict that the $(i+1)$--th firing of transition $t$ must be scheduled after the $i$--th execution, unless the $(i+1)$--th execution is not scheduled before simulation termination. The value of $M$ in constraints (C3) and (C4) can be set to $K$.

\begin{eqnarray}
	\sum_{k\in \mathbb{K}} x^{t}_{i+1,k} - \sum_{k\in \mathbb{K}} x^{t}_{i,k} \le 0&  \forall\ t\in \mathbb{T},i\in \mathbb{I}^{t}&(C1)\nonumber\\
	\sum_{k\in \mathbb{K}} w^{t}_{i,k}-\sum_{k\in \mathbb{K}} x^{t}_{i,k} \le 0&  \forall\ t\in \mathbb{T},i\in \mathbb{I}^{t}&(C2)\nonumber\\
	\sum_{k\in \mathbb{K}} kw^{t}_{i,k} - \sum_{k\in \mathbb{K}} kx^{t}_{i,k} \ge M(\sum_{k\in \mathbb{K}} w^{t}_{i,k}-1) &\forall\ t\in \mathbb{T}, i\in \mathbb{I}^{t}&(C3) \nonumber\\
	\sum_{k\in \mathbb{K}} kx^{t}_{i+1,k} - \sum_{k\in \mathbb{K}} kx^{t}_{i,k} \ge 1+M(\sum_{k\in \mathbb{K}} x^{t}_{i+1,k}-1)&  \forall\ t\in \mathbb{T},i\in \mathbb{I}^{t}&(C4)\nonumber
\end{eqnarray}

@@

The tokens in place $p$ are changed from $m^{p}_{k}$ to $m^{p}_{k+1}$ in iteration $k$, as in constraints (D1) and in lines 16 to 18 in Algorithm 1. The transitions $t$ that start in %fire 
firing in iteration $k$ absorbs $A^{p,t}$ tokens from the precedent places $p$, and the firing that finishes in iteration $k$ releases tokens to the successive places.

\begin{eqnarray}
	m^p_{k+1} = m^p_k - \sum_{t\in \mathbb{T}} W^{p,t} \sum_{i\in \mathbb{I}^{t}} x^t_{i,k} +  \sum_{t\in \mathbb{T}} W^{t,p} \sum_{i\in \mathbb{I}^{t}} w^t_{i,k} &\forall p\in \mathbb{P}, k\in \mathbb{K}& (D1) \nonumber
\end{eqnarray}

Equation (D1) can also be reformulated as (D2).
\begin{eqnarray}
	m^p_{k+1} = m^p_0 - \sum_{t\in \mathbb{T}} W^{p,t} \sum_{k^{'}=0}^k \sum_{i\in \mathbb{I}^{t}} x^t_{i,k^{'}} +  \sum_{t\in \mathbb{T}} W^{t,p} \sum_{k^{'}=0}^k \sum_{i\in \mathbb{I}^{t}} w^t_{i,k^{'}} &\forall p\in \mathbb{P}, k\in \mathbb{K}& (D2) \nonumber
\end{eqnarray}

\subsection{Objective function}\label{sec:TPN_obj}
With the constraints defined in the previous sections, there is a unique feasible solution in terms of transition firing times and the %history of simulation clock, 
sequence of simulation clock values, i.e., $e^{t,s}_i$, $e^{t,f}_{i}$ and $\mathcal{E}_k$. Thus, the objective function can be any function of %those
these variables, such as, average system time or average waiting time in queueing systems. Notice that multiple feasible solutions may appear in terms of binary variables, since the sequence of events with identical execution time is not uniquely defined.

@@

\section{Indicative examples} \label{sec:example}
In this section, the $G/G/m$ is used as a example to generate MPR from TPN. Both the proposed and the state-of-the-art formulation are both presented. Finally, an extended model to optimize the $G/G/m$ queue is stated.

\subsection{MPR of G/G/m queue}\label{sec:MPR_ggm}
A $G/G/m$ queue is composed by $m$ parallel identical servers. Customers arrive at the queue following an general arrival process. One of the identical idle servers processes customers in the queue one by one, and the processing time is generally distributed. After being processed, the customer will be released from the system immediately, and the server becomes idle again. In Figure \ref{fig:tpn_GG2}, the TPN of $G/G/2$ is presented. A token in place $p_{arr}$ shows that it is possible to schedule the arrival of next customer, and the firing time of transition $t_{arr}$ is equal to the inter-arrival time. At the end of firing $t_{arr}$, a customer arrives, and it is possible to schedule the next arrival, thus, it releases one token to $p_{arr}$. Furthermore, an arrival will increase the length of the queue by 1, which is represented by the number of tokens in place $p_{queue}$. Two tokens in place $p_{idle}$ shows that two servers are idle. The firing of transition $t_{process}$ represents the processing of a customer, and it is enabled if and only if at least one server is idle and there is a customer in the queue. Once the process of the customer finishes, i.e., the end of firing $t_{process}$, a server is released, and a token is sent to place $p_{idle}$.

@@

As mentioned in Section \ref{sec:MPR_start_Firing}, the transitions with non-zero firing times are expanded before implementing the MPR generation. In the case of G/G/2 queue, transition $t_{process}$ needs to be expanded to $\tilde{t}_{process}$, $\bar{t}_{process}$ and $p_{process}$. Transition $t_{arr}$ does not need expansion since it is self-limiting. The resulting expanded TPN is shown in Figure \ref{fig:tpn_GG2_expansion}.

@@

The sets $\mathbb{T},\ \mathbb{K},\ \mathbb{I}^t$ are specified as follows. Set $\mathbb{T}$ is composed of three transitions $\{t_{arr},\tilde{t}_{process}, \bar{t}_{process}\}$ as shown in Figure \ref{fig:tpn_GG2_expansion}. If $n$ jobs' arrival, processing and departure are simulated, the three transitions are fired for $n$ times each, and set $\mathbb{I}^t$ is equal to $\{1,2,..., n\}$. There are $3n$ transition firings in the simulation execution and the set $\mathbb{K}$ is equal to $\{0,1,..., 3n-1\}$. The initial markings $m^{p_{arr}}_0,\ m^{p_{queue}}_0, m^{p_{process}}_0, m^{p_{idle}}$ are specified as 1, 0, 0, 2. Once the TPN with the initial markings and simulation length are provided, the MPR composed of (A1) to (A7), (B1) to (B7), (C1) to (C4) and (D1) can be generated.

\begin{eqnarray}
	\mathbb{T} = \{t_{arr},\tilde{t}_{proc}, \bar{t}_{proc}\} \nonumber\\
	\mathbb{K} = \{0,1,..., 3n-1\} \nonumber\\
	\mathbb{I}^t=\{1,2,..., n\} \nonumber
\end{eqnarray}

Table \ref{tab:sim_GGM} shows the first ten iteration of a simulation run of a G/G/m queue.

In iteration 0, the system is in its initial state with markings (1,0,0,2), and transition $t_{arr}$ can be fired, equivalently, binary variables $z^{arr}_0$ and $x^{arr}_{1,0}$ are equal to $1$ in the solution of the MPR. The time to start fire transition $t_{arr}$ is equal to $0$, and the variable $e^{arr,s}_{1}$ takes value $0$ in the MPR solution. The fire will finish at time 2.3, and variable $e^{arr,f}_{1}$ takes value 2.3 equivalently. That firing is also the firing to finish in this iteration, and $w^{arr}_{1,0}$ is equal to 1. After that, the markings are changed to (1,1,0,2) and this iteration ends.
